---
title: "Pitch Metrics Leaderboard — 2025"
author: "Carl Foust"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)
library(xgboost)
library(RPostgres)
library(DT)
library(knitr)

source("../utils.R")
source("data_prep.R")

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
set.seed(1015)

MODELS_DIR <- "~/Code/Stuff-Plus/models/"
MIN_PITCHES <- 200   # minimum pitches to appear on leaderboard

# Linear run value weights
RV_WHIFF <- -0.25   # swinging strike
RV_FOUL  <- -0.07   # foul ball
RV_CS    <- -0.12   # called strike
RV_HBP   <-  0.37   # hit by pitch
# BIP run values by launch speed-angle category (classes 0–5):
# 0=Weak, 1=Topped, 2=Under, 3=Flare/Burner, 4=Solid, 5=Barrel
RV_BIP   <- c(-0.14, -0.09, -0.03, 0.09, 0.15, 0.28)
```

```{r helpers}
# Find and load the most-recently-trained RDS workflow for a given model name
load_model <- function(name, dir = MODELS_DIR) {
  pattern <- paste0("^", name, "_XGBoost_.*\\.rds$")
  files   <- list.files(dir, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) {
    warning(paste0("Model not found: ", name))
    return(NULL)
  }
  readRDS(files[order(file.mtime(files), decreasing = TRUE)][1])
}

# Binary prediction → P(positive class)
pred_binary <- function(model, data, features) {
  if (is.null(model) || nrow(data) == 0) return(rep(NA_real_, nrow(data)))
  tryCatch(
    predict(model, data |> select(all_of(features)), type = "prob")$.pred_1,
    error = function(e) { warning(e$message); rep(NA_real_, nrow(data)) }
  )
}

# Multiclass BIP prediction → expected run value (dot product with RV_BIP)
pred_bip_rv <- function(model, data, features) {
  if (is.null(model) || nrow(data) == 0) return(rep(NA_real_, nrow(data)))
  tryCatch({
    probs <- predict(model, data |> select(all_of(features)), type = "prob")
    as.numeric(as.matrix(probs) %*% RV_BIP)
  }, error = function(e) { warning(e$message); rep(NA_real_, nrow(data)) })
}

# Scale a per-pitcher average RV to a Plus metric (100 = league avg, higher = better)
to_plus <- function(pitcher_rv) {
  league_mean <- mean(pitcher_rv, na.rm = TRUE)
  league_sd   <- sd(pitcher_rv,   na.rm = TRUE)
  round(100 + (league_mean - pitcher_rv) / league_sd * 15, 1)
}
```

```{r feature-definitions}
# Feature sets mirror the training modelPrep column selections exactly
FF_STUFF <- c("is_lhh", "release_speed", "release_spin_rate", "release_extension",
              "release_pos_x", "release_pos_z", "a_vert", "diff_exp_avert",
              "a_horz", "diff_exp_ahorz", "spin_axis", "arm_angle", "xVAA", "xHAA")

BB_STUFF <- c("is_lhh", "release_speed", "release_spin_rate", "release_extension",
              "release_pos_x", "release_pos_z", "a_vert", "a_horz",
              "spin_axis", "arm_angle", "xVAA", "xHAA",
              "rel_speed_fb_diff", "a_horz_fb_diff", "a_vert_fb_diff")

OS_STUFF <- c("is_lhh", "release_speed", "release_spin_rate", "release_extension",
              "release_pos_x", "release_pos_z", "a_vert", "a_horz",
              "spin_axis", "arm_angle",
              "rel_speed_fb_diff", "a_horz_fb_diff", "a_vert_fb_diff")

LOC_FEAT  <- c("is_lhh", "plate_x", "plate_z", "sz_top", "sz_bot", "dist_from_mid")

FF_QUAL   <- c(FF_STUFF, "plate_x", "plate_z", "sz_top", "sz_bot", "dist_from_mid")
BB_QUAL   <- c(BB_STUFF, "plate_x", "plate_z", "sz_top", "sz_bot", "dist_from_mid")
OS_QUAL   <- c(OS_STUFF, "plate_x", "plate_z", "sz_top", "sz_bot", "dist_from_mid")
```

---

## Model Loading

```{r load-secondary-models}
# Secondary models used to generate input features for Stuff+ and Quality+
.vaa_model        <- load_model_xgb("exp_vaa_XGBoost_2025-12-24_12-28.json")
.haa_model        <- load_model_xgb("exp_haa_XGBoost_2025-12-24_14-31.json")
.arm_angle_model  <- load_model_xgb("exp_arm_angle_XGBoost_2025-12-24_16-43.json")
.exp_ff_avert_model  <- load_model_xgb("exp_ff_az_XGBoost_2026-01-06_00-04.json")
.exp_ff_ahorz_model  <- load_model_xgb("exp_ff_ahorz_XGBoost_2026-01-19_21-30.json")
.exp_si_avert_model  <- load_model_xgb("exp_si_avert_XGBoost_2026-02-16_23-34.json")
.exp_si_ahorz_model  <- load_model_xgb("exp_si_ahorz_XGBoost_2026-02-17_00-02.json")
.exp_ct_avert_model  <- load_model_xgb("exp_ct_avert_XGBoost_2026-02-16_22-01.json")
.exp_ct_ahorz_model  <- load_model_xgb("exp_ct_ahorz_XGBoost_2026-02-16_22-34.json")
```

```{r load-stuff-models}
stuff_models <- list(
  ff = list(
    rhp = list(whiff = load_model("rhp_ff_whiff"), foul = load_model("rhp_ff_foul"), bip = load_model("rhp_ff_bip")),
    lhp = list(whiff = load_model("lhp_ff_whiff"), foul = load_model("lhp_ff_foul"), bip = load_model("lhp_ff_bip"))
  ),
  bb = list(
    rhp = list(whiff = load_model("rhp_bb_whiff"), foul = load_model("rhp_bb_foul"), bip = load_model("rhp_bb_bip")),
    lhp = list(whiff = load_model("lhp_bb_whiff"), foul = load_model("lhp_bb_foul"), bip = load_model("lhp_bb_bip"))
  ),
  os = list(
    rhp = list(whiff = load_model("rhp_os_whiff"), foul = load_model("rhp_os_foul"), bip = load_model("rhp_os_bip")),
    lhp = list(whiff = load_model("lhp_os_whiff"), foul = load_model("lhp_os_foul"), bip = load_model("lhp_os_bip"))
  )
)
```

```{r load-location-models}
loc_models <- list(
  ff = list(
    rhp = list(whiff = load_model("rhp_ff_loc_whiff"), foul = load_model("rhp_ff_loc_foul"),
               bip   = load_model("rhp_ff_loc_bip"),   cs   = load_model("rhp_ff_loc_cs"),
               hbp   = load_model("rhp_ff_loc_hbp")),
    lhp = list(whiff = load_model("lhp_ff_loc_whiff"), foul = load_model("lhp_ff_loc_foul"),
               bip   = load_model("lhp_ff_loc_bip"),   cs   = load_model("lhp_ff_loc_cs"),
               hbp   = load_model("lhp_ff_loc_hbp"))
  ),
  bb = list(
    rhp = list(whiff = load_model("rhp_bb_loc_whiff"), foul = load_model("rhp_bb_loc_foul"),
               bip   = load_model("rhp_bb_loc_bip"),   cs   = load_model("rhp_bb_loc_cs"),
               hbp   = load_model("rhp_bb_loc_hbp")),
    lhp = list(whiff = load_model("lhp_bb_loc_whiff"), foul = load_model("lhp_bb_loc_foul"),
               bip   = load_model("lhp_bb_loc_bip"),   cs   = load_model("lhp_bb_loc_cs"),
               hbp   = load_model("lhp_bb_loc_hbp"))
  ),
  os = list(
    rhp = list(whiff = load_model("rhp_os_loc_whiff"), foul = load_model("rhp_os_loc_foul"),
               bip   = load_model("rhp_os_loc_bip"),   cs   = load_model("rhp_os_loc_cs"),
               hbp   = load_model("rhp_os_loc_hbp")),
    lhp = list(whiff = load_model("lhp_os_loc_whiff"), foul = load_model("lhp_os_loc_foul"),
               bip   = load_model("lhp_os_loc_bip"),   cs   = load_model("lhp_os_loc_cs"),
               hbp   = load_model("lhp_os_loc_hbp"))
  )
)
```

```{r load-quality-models}
qual_models <- list(
  ff = list(
    rhp = list(whiff = load_model("rhp_ff_qual_whiff"), foul = load_model("rhp_ff_qual_foul"),
               bip   = load_model("rhp_ff_qual_bip"),   cs   = load_model("rhp_ff_qual_cs"),
               hbp   = load_model("rhp_ff_qual_hbp")),
    lhp = list(whiff = load_model("lhp_ff_qual_whiff"), foul = load_model("lhp_ff_qual_foul"),
               bip   = load_model("lhp_ff_qual_bip"),   cs   = load_model("lhp_ff_qual_cs"),
               hbp   = load_model("lhp_ff_qual_hbp"))
  ),
  bb = list(
    rhp = list(whiff = load_model("rhp_bb_qual_whiff"), foul = load_model("rhp_bb_qual_foul"),
               bip   = load_model("rhp_bb_qual_bip"),   cs   = load_model("rhp_bb_qual_cs"),
               hbp   = load_model("rhp_bb_qual_hbp")),
    lhp = list(whiff = load_model("lhp_bb_qual_whiff"), foul = load_model("lhp_bb_qual_foul"),
               bip   = load_model("lhp_bb_qual_bip"),   cs   = load_model("lhp_bb_qual_cs"),
               hbp   = load_model("lhp_bb_qual_hbp"))
  ),
  os = list(
    rhp = list(whiff = load_model("rhp_os_qual_whiff"), foul = load_model("rhp_os_qual_foul"),
               bip   = load_model("rhp_os_qual_bip"),   cs   = load_model("rhp_os_qual_cs"),
               hbp   = load_model("rhp_os_qual_hbp")),
    lhp = list(whiff = load_model("lhp_os_qual_whiff"), foul = load_model("lhp_os_qual_foul"),
               bip   = load_model("lhp_os_qual_bip"),   cs   = load_model("lhp_os_qual_cs"),
               hbp   = load_model("lhp_os_qual_hbp"))
  )
)
```

---

## Data Preparation

```{r db-pull}
con <- dbConnect(Postgres(),
                 user = "postgres",
                 host = "localhost",
                 db   = "baseball-research")

sql_2025 <- "SELECT * FROM statcast_pitching_dev WHERE game_year = 2025 AND game_type = 'R'"
data_2025 <- dbGetQuery(con, sql_2025)
dbDisconnect(con)
```

```{r process-data}
processed <- create_advance_metrics(data_2025)
```

```{r secondary-preds}
vaa_preds          <- generate_model_preds(.vaa_model,       processed, vaa_features,       "xVAA")
haa_preds          <- generate_model_preds(.haa_model,       processed, haa_features,       "xHAA")
arm_angle_preds    <- generate_model_preds(.arm_angle_model,  processed, arm_angle_features, "exp_arm_angle")
exp_ff_avert_preds <- generate_model_preds(.exp_ff_avert_model, processed, exp_ff_features,   "exp_ff_avert")
exp_ff_ahorz_preds <- generate_model_preds(.exp_ff_ahorz_model, processed, exp_ff_x_features, "exp_ff_ahorz")
exp_si_avert_preds <- generate_model_preds(.exp_si_avert_model, processed, exp_ff_features,   "exp_si_avert")
exp_si_ahorz_preds <- generate_model_preds(.exp_si_ahorz_model, processed, exp_ff_x_features, "exp_si_ahorz")
exp_ct_avert_preds <- generate_model_preds(.exp_ct_avert_model, processed, exp_ff_features,   "exp_ct_avert")
exp_ct_ahorz_preds <- generate_model_preds(.exp_ct_ahorz_model, processed, exp_ff_x_features, "exp_ct_ahorz")

pitching <- bind_cols(
  processed,
  vaa_preds, haa_preds, arm_angle_preds,
  exp_ff_avert_preds, exp_ff_ahorz_preds,
  exp_si_avert_preds, exp_si_ahorz_preds,
  exp_ct_avert_preds, exp_ct_ahorz_preds
) |>
  mutate(
    arm_angle = ifelse(is.na(arm_angle), exp_arm_angle, arm_angle),
    # Fastball-specific expected acceleration and deviation columns
    exp_ahorz = case_when(pitch_name == "4-Seam Fastball" ~ exp_ff_ahorz,
                          pitch_name == "Sinker"          ~ exp_si_ahorz,
                          pitch_name == "Cutter"          ~ exp_ct_ahorz,
                          TRUE ~ NA_real_),
    exp_avert = case_when(pitch_name == "4-Seam Fastball" ~ exp_ff_avert,
                          pitch_name == "Sinker"          ~ exp_si_avert,
                          pitch_name == "Cutter"          ~ exp_ct_avert,
                          TRUE ~ NA_real_),
    diff_exp_ahorz = a_horz - exp_ahorz,
    diff_exp_avert = a_vert - exp_avert
  )
```

---

## Generating Predictions

```{r predict-function}
# Predict all three metrics for one pitch group × handedness slice
predict_slice <- function(data_slice, hand_models_stuff, hand_models_loc, hand_models_qual,
                          feat_stuff, feat_loc, feat_qual) {
  if (nrow(data_slice) == 0) return(NULL)

  # Stuff+ run value (swing-conditional quality)
  rv_stuff <- pred_binary(hand_models_stuff$whiff, data_slice, feat_stuff) * RV_WHIFF +
              pred_binary(hand_models_stuff$foul,  data_slice, feat_stuff) * RV_FOUL  +
              pred_bip_rv(hand_models_stuff$bip,   data_slice, feat_stuff)

  # Location+ run value
  rv_loc_swing <- pred_binary(hand_models_loc$whiff, data_slice, feat_loc) * RV_WHIFF +
                  pred_binary(hand_models_loc$foul,  data_slice, feat_loc) * RV_FOUL  +
                  pred_bip_rv(hand_models_loc$bip,   data_slice, feat_loc)
  rv_loc_all   <- pred_binary(hand_models_loc$cs,  data_slice, feat_loc) * RV_CS  +
                  pred_binary(hand_models_loc$hbp,  data_slice, feat_loc) * RV_HBP
  rv_loc       <- rv_loc_swing + rv_loc_all

  # Quality+ run value
  rv_qual_swing <- pred_binary(hand_models_qual$whiff, data_slice, feat_qual) * RV_WHIFF +
                   pred_binary(hand_models_qual$foul,  data_slice, feat_qual) * RV_FOUL  +
                   pred_bip_rv(hand_models_qual$bip,   data_slice, feat_qual)
  rv_qual_all   <- pred_binary(hand_models_qual$cs,  data_slice, feat_qual) * RV_CS  +
                   pred_binary(hand_models_qual$hbp,  data_slice, feat_qual) * RV_HBP
  rv_qual       <- rv_qual_swing + rv_qual_all

  data_slice |>
    select(pitcher, pitch_name, pitch_group, p_throws, game_date) |>
    mutate(rv_stuff = rv_stuff, rv_loc = rv_loc, rv_qual = rv_qual)
}
```

```{r run-predictions}
groups <- list(
  list(group = "Fastball",    key = "ff", feat_stuff = FF_STUFF, feat_qual = FF_QUAL),
  list(group = "BreakingBall",key = "bb", feat_stuff = BB_STUFF, feat_qual = BB_QUAL),
  list(group = "Offspeed",    key = "os", feat_stuff = OS_STUFF, feat_qual = OS_QUAL)
)

all_preds <- map_dfr(groups, function(g) {
  d <- pitching |> filter(pitch_group == g$group)

  bind_rows(
    predict_slice(d |> filter(p_throws == 1),
                  stuff_models[[g$key]]$rhp, loc_models[[g$key]]$rhp, qual_models[[g$key]]$rhp,
                  g$feat_stuff, LOC_FEAT, g$feat_qual),
    predict_slice(d |> filter(p_throws == 0),
                  stuff_models[[g$key]]$lhp, loc_models[[g$key]]$lhp, qual_models[[g$key]]$lhp,
                  g$feat_stuff, LOC_FEAT, g$feat_qual)
  )
})
```

---

## Leaderboard

```{r build-leaderboard}
# Join player names
player_names <- data_2025 |>
  distinct(pitcher, player_name)

leaderboard_raw <- all_preds |>
  group_by(pitcher, pitch_group) |>
  summarise(
    n         = n(),
    rv_stuff  = mean(rv_stuff, na.rm = TRUE),
    rv_loc    = mean(rv_loc,   na.rm = TRUE),
    rv_qual   = mean(rv_qual,  na.rm = TRUE),
    .groups   = "drop"
  ) |>
  filter(n >= MIN_PITCHES) |>
  left_join(player_names, by = "pitcher")

# Overall (all pitch types combined)
leaderboard_overall <- all_preds |>
  group_by(pitcher) |>
  summarise(
    n         = n(),
    rv_stuff  = mean(rv_stuff, na.rm = TRUE),
    rv_loc    = mean(rv_loc,   na.rm = TRUE),
    rv_qual   = mean(rv_qual,  na.rm = TRUE),
    .groups   = "drop"
  ) |>
  filter(n >= MIN_PITCHES) |>
  left_join(player_names, by = "pitcher") |>
  mutate(
    stuff_plus    = to_plus(rv_stuff),
    location_plus = to_plus(rv_loc),
    quality_plus  = to_plus(rv_qual)
  ) |>
  arrange(desc(stuff_plus)) |>
  select(player_name, n, stuff_plus, location_plus, quality_plus)

# Per-group leaderboards
make_group_leaderboard <- function(df, group_name) {
  df |>
    filter(pitch_group == group_name) |>
    mutate(
      stuff_plus    = to_plus(rv_stuff),
      location_plus = to_plus(rv_loc),
      quality_plus  = to_plus(rv_qual)
    ) |>
    arrange(desc(stuff_plus)) |>
    select(player_name, n, stuff_plus, location_plus, quality_plus)
}

lb_ff <- make_group_leaderboard(leaderboard_raw, "Fastball")
lb_bb <- make_group_leaderboard(leaderboard_raw, "BreakingBall")
lb_os <- make_group_leaderboard(leaderboard_raw, "Offspeed")
```

### Overall (All Pitches)

```{r table-overall}
datatable(
  leaderboard_overall |> rename(Pitcher = player_name, Pitches = n,
                                 `Stuff+` = stuff_plus, `Location+` = location_plus,
                                 `Quality+` = quality_plus),
  options  = list(pageLength = 25, order = list(list(2, "desc"))),
  rownames = FALSE
) |>
  formatRound(columns = c("Stuff+", "Location+", "Quality+"), digits = 1)
```

### Fastballs

```{r table-ff}
datatable(
  lb_ff |> rename(Pitcher = player_name, Pitches = n,
                   `Stuff+` = stuff_plus, `Location+` = location_plus,
                   `Quality+` = quality_plus),
  options  = list(pageLength = 25, order = list(list(2, "desc"))),
  rownames = FALSE
) |>
  formatRound(columns = c("Stuff+", "Location+", "Quality+"), digits = 1)
```

### Breaking Balls

```{r table-bb}
datatable(
  lb_bb |> rename(Pitcher = player_name, Pitches = n,
                   `Stuff+` = stuff_plus, `Location+` = location_plus,
                   `Quality+` = quality_plus),
  options  = list(pageLength = 25, order = list(list(2, "desc"))),
  rownames = FALSE
) |>
  formatRound(columns = c("Stuff+", "Location+", "Quality+"), digits = 1)
```

### Off-Speed

```{r table-os}
datatable(
  lb_os |> rename(Pitcher = player_name, Pitches = n,
                   `Stuff+` = stuff_plus, `Location+` = location_plus,
                   `Quality+` = quality_plus),
  options  = list(pageLength = 25, order = list(list(2, "desc"))),
  rownames = FALSE
) |>
  formatRound(columns = c("Stuff+", "Location+", "Quality+"), digits = 1)
```

---

## Metric Definitions

| Metric | What it measures | Features used |
|---|---|---|
| **Stuff+** | Expected swing-outcome quality based on pitch physical characteristics | Velocity, movement, spin, approach angles |
| **Location+** | Expected outcome quality based purely on pitch placement and batter strike zone | Plate coordinates, strike zone dimensions |
| **Quality+** | Combined expected outcome quality (stuff + location) | All Stuff+ features plus plate coordinates |

**Scaling:** 100 = league average, 115 ≈ 1 SD above average (better for pitcher), 85 ≈ 1 SD below. All metrics include Whiff, Foul, BIP quality (swing-conditional), Called Strike, and HBP.
