---
title: "Player-Level SHAP Analysis"
author: "Carl Foust"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Data Manipulation
library(tidyverse)

# Modeling
library(tidymodels)
library(xgboost)
library(shapviz)

# SQL
library(RMySQL)

wd <- getwd()
set.seed(1015)
```

## Parameters

Edit these values to select the player and model you want to analyze.

```{r parameters}
# --- USER CONFIG ---
target_player <- "Gerrit Cole"   # player_name as it appears in Statcast data
target_hand   <- "RHP"           # "RHP" or "LHP"
pitch_type    <- "Fastball"      # "Fastball", "BreakingBall", "OffSpeed"
outcome_type  <- "Whiff"         # "Whiff", "Foul", "BIP"
model_dir     <- "../models"     # path to the directory containing saved RDS models

# Path to approach angle models (used for xVAA / xHAA feature calculation)
vaa_model_path <- file.path(model_dir, "VAA.rds")
haa_model_path <- file.path(model_dir, "HAA.rds")

# How many pitches to use for SHAP (NULL = all; set a number to sample for speed)
shap_sample_n <- NULL

# Output directory for saved plots
output_dir <- file.path(wd, "output")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

## Load Approach Angle Models

```{r load-angle-models}
.VAA_Model <- readRDS(vaa_model_path)
.HAA_Model <- readRDS(haa_model_path)
```

## Load & Preprocess Data

Full feature engineering pipeline, identical to the primary model training scripts.
Player name is retained throughout so we can filter after preprocessing.

```{r load-data}
con <- dbConnect(MySQL(),
                 user = "root",
                 host = "localhost",
                 db   = "baseball-research")

# Use all available data (train + test) for player-level analysis
sql <- "SELECT * FROM statcast_pitching WHERE game_year >= 2021 AND game_type = 'R'"
raw_data <- dbGetQuery(con, sql)
dbDisconnect(con)
```

```{r feature-engineering}
pitching_pbp <- raw_data |>
  select(player_name, pitcher, pitch_name, at_bat_number,
         pitch_number, description, launch_speed_angle, plate_x, plate_z,
         release_speed, release_spin_rate, release_extension, spin_axis,
         release_pos_x, release_pos_z, pfx_x, pfx_z,
         p_throws, stand, game_year, game_date, vx0,
         vy0, vz0, ax, ay, az, arm_angle) |>
  drop_na(!launch_speed_angle)

# Compute approach angles and acceleration components
pitching_processed <- pitching_pbp |>
  mutate(vy_f = -sqrt((vy0^2 - (2 * ay * (50 - (17/12))))),
         t    = (vy_f - vy0) / ay,
         vz_f = vz0 + (az * t),
         vx_f = vx0 + (ax * t),
         VAA  = -atan((vz_f / vy_f)) * (180 / pi),
         HAA  = -atan((vx_f / vy_f)) * (180 / pi),
         v_mag    = sqrt(vx0^2 + vy0^2 + vz0^2),
         tang_x   = vx0 / v_mag,
         tang_y   = vy0 / v_mag,
         tang_z   = vz0 / v_mag,
         handedness_factor = if_else(p_throws == "R", 1, -1),
         temp_x = 0, temp_y = 0, temp_z = 1,
         horz_x_raw = temp_y * tang_z - temp_z * tang_y,
         horz_y_raw = temp_z * tang_x - temp_x * tang_z,
         horz_z_raw = temp_x * tang_y - temp_y * tang_x,
         horz_mag   = sqrt(horz_x_raw^2 + horz_y_raw^2 + horz_z_raw^2),
         horz_x = horz_x_raw / horz_mag,
         horz_y = horz_y_raw / horz_mag,
         horz_z = horz_z_raw / horz_mag,
         vert_x = tang_y * horz_z - tang_z * horz_y,
         vert_y = tang_z * horz_x - tang_x * horz_z,
         vert_z = tang_x * horz_y - tang_y * horz_x,
         a_tang = ax * tang_x + ay * tang_y + az * tang_z,
         a_horz = ax * horz_x + ay * horz_y + az * horz_z,
         a_vert = ax * vert_x + ay * vert_y + az * vert_z) |>
  select(-c(vx0:az, vy_f:vx_f, v_mag:a_tang))

pitching_processed <- pitching_processed |>
  mutate(pfx_x      = pfx_x * 12,
         pfx_z      = pfx_z * 12,
         stand      = as.numeric(factor(stand)) - 1,
         plate_x_og = plate_x,
         plate_z_og = plate_z,
         plate_x    = 0,
         plate_z    = 2.5)

# Pitch group classification
pitching_processed <- pitching_processed |>
  mutate(PitchGroup = case_when(
    pitch_name %in% c("4-Seam Fastball", "Sinker", "Cutter")                          ~ "Fastball",
    pitch_name %in% c("Slider", "Sweeper", "Curveball", "Slurve",
                      "Knuckle Curve", "Slow Curve", "Screwball")                      ~ "BreakingBall",
    pitch_name %in% c("Changeup", "Split-Finger", "Forkball")                         ~ "Offspeed",
    pitch_name %in% c("Eephus", "Knuckleball", "Other")                               ~ "Other",
    TRUE ~ NA
  ))

# xVAA / xHAA predictions
VAA_preds <- predict(.VAA_Model, pitching_processed) |> rename(xVAA = .pred)
HAA_preds <- predict(.HAA_Model, pitching_processed) |> rename(xHAA = .pred)

pitching_processed <- bind_cols(pitching_processed, VAA_preds, HAA_preds) |>
  select(player_name:pitch_name, PitchGroup, at_bat_number:stand,
         arm_angle:VAA, xVAA, HAA, xHAA, a_horz:plate_z_og, game_date, game_year)

# Fastball averages for relative features
FB_avg <- pitching_processed |>
  filter(pitch_name %in% c("4-Seam Fastball", "Sinker")) |>
  group_by(player_name, pitcher) |>
  summarise(n       = n(),
            FBvelo  = mean(release_speed, na.rm = TRUE),
            FBvbreak = mean(pfx_z, na.rm = TRUE),
            FBhbreak = mean(pfx_x, na.rm = TRUE),
            .groups = "drop")

# Cutter reclassification
Cutter_data <- pitching_processed |>
  filter(pitch_name == "Cutter") |>
  group_by(player_name, pitcher) |>
  summarise(n = n(), CTvelo = mean(release_speed, na.rm = TRUE),
            CTvert = mean(pfx_z, na.rm = TRUE), .groups = "drop")

Cutter_labeled <- Cutter_data |>
  left_join(FB_avg, by = c("player_name", "pitcher")) |>
  mutate(VeloRatio  = CTvelo / FBvelo,
         pitch_name = "Cutter",
         new_group  = case_when(
           is.na(FBvelo) & CTvert > 5  ~ "Fastball",
           is.na(FBvelo)               ~ "BreakingBall",
           VeloRatio >= 0.95           ~ "Fastball",
           CTvert <= 5                 ~ "BreakingBall",
           VeloRatio <= 0.935          ~ "BreakingBall",
           TRUE                        ~ "Fastball"
         )) |>
  select(player_name, pitcher, pitch_name, new_group)

pitching_processed <- pitching_processed |>
  left_join(Cutter_labeled, by = c("player_name", "pitcher", "pitch_name")) |>
  mutate(PitchGroup = if_else(pitch_name == "Cutter", new_group, PitchGroup)) |>
  select(-new_group)

# Relative features
pitching_processed <- pitching_processed |>
  left_join(FB_avg |> select(-n), by = c("player_name", "pitcher")) |>
  mutate(RelSpeedDiff    = release_speed - FBvelo,
         Horz_Accel_Diff = a_horz - FBhbreak,
         Vert_Accel_Diff = a_vert - FBvbreak,
         launch_speed_angle = ifelse(is.na(launch_speed_angle), 0, launch_speed_angle)) |>
  na.omit()

# Target variables
Swing_desc <- c("hit_into_play", "foul", "foul_tip", "swinging_strike",
                "swinging_strike_blocked", "foul_bunt", "missed_bunt", "bunt_foul_tip")
Whiff_desc <- c("swinging_strike", "swinging_strike_blocked", "foul_tip", "bunt_foul_tip")
Foul_desc  <- c("foul", "foul_bunt")

pitching_processed <- pitching_processed |>
  mutate(Swing = case_when(description %in% Swing_desc ~ TRUE, TRUE ~ 0),
         Whiff = case_when(description %in% Whiff_desc ~ TRUE, TRUE ~ 0),
         Foul  = case_when(description %in% Foul_desc  ~ TRUE, TRUE ~ 0),
         BIP   = launch_speed_angle)
```

## Load Model

```{r load-model}
model_files <- list.files(
  path    = file.path(model_dir, target_hand, outcome_type),
  pattern = paste0("^", pitch_type, "_.*\\.rds$"),
  full.names = TRUE
)

if (length(model_files) == 0) {
  stop("No model found for: ", target_hand, " / ", pitch_type, " / ", outcome_type,
       "\nLooked in: ", file.path(model_dir, target_hand, outcome_type))
}

model_path  <- model_files[which.max(file.mtime(model_files))]
final_model <- readRDS(model_path)
message("Loaded model: ", basename(model_path))
```

## Prepare Player Data

```{r player-data}
feature_cols <- c("release_speed", "release_spin_rate", "release_extension",
                  "release_pos_x", "release_pos_z", "a_horz", "a_vert",
                  "spin_axis", "arm_angle", "stand", "xVAA", "xHAA")

p_throws_val <- if_else(target_hand == "RHP", "R", "L")

player_data <- pitching_processed |>
  filter(player_name == target_player,
         p_throws    == p_throws_val,
         PitchGroup  == pitch_type)

if (outcome_type == "BIP") {
  player_data <- player_data |> filter(BIP != 0)
}

if (nrow(player_data) == 0) {
  stop("No pitches found for player: ", target_player,
       " (", target_hand, " | ", pitch_type, ")")
}

# Optional: sample for speed
if (!is.null(shap_sample_n) && nrow(player_data) > shap_sample_n) {
  set.seed(1015)
  player_data <- player_data |> slice_sample(n = shap_sample_n)
}

X_player <- player_data |> select(all_of(feature_cols)) |> as.matrix()

message(nrow(X_player), " pitches for ", target_player)
```

## Compute SHAP Values

```{r shap-compute}
xgb_engine <- extract_fit_engine(final_model)
shap_obj   <- shapviz(xgb_engine, X_pred = X_player)
```

## SHAP Plots

### Feature Importance — Beeswarm

Each dot is one pitch. Color indicates the feature value (red = high, blue = low).
Horizontal position shows the SHAP value — how much that feature pushed the prediction
up (right) or down (left).

```{r shap-beeswarm, fig.width=10, fig.height=7}
p_beeswarm <- sv_importance(shap_obj, kind = "beeswarm", max_display = 12) +
  labs(title = paste0(target_player, "  |  ", pitch_type, " ", outcome_type),
       subtitle = paste0("SHAP Feature Importance  (n = ", nrow(X_player), " pitches)"))

p_beeswarm

ggsave(filename = file.path(output_dir,
         paste0(gsub(" ", "_", target_player), "_", target_hand, "_",
                pitch_type, "_", outcome_type, "_SHAP_beeswarm.png")),
       plot = p_beeswarm, width = 10, height = 7, dpi = 300)
```

### Feature Importance — Bar (Mean |SHAP|)

Average absolute SHAP value per feature — a simple ranking of overall influence.

```{r shap-bar, fig.width=9, fig.height=6}
p_bar <- sv_importance(shap_obj, kind = "bar", max_display = 12) +
  labs(title = paste0(target_player, "  |  ", pitch_type, " ", outcome_type),
       subtitle = "Mean |SHAP| — Average feature influence across all pitches")

p_bar

ggsave(filename = file.path(output_dir,
         paste0(gsub(" ", "_", target_player), "_", target_hand, "_",
                pitch_type, "_", outcome_type, "_SHAP_bar.png")),
       plot = p_bar, width = 9, height = 6, dpi = 300)
```

### Waterfall — Single Pitch Breakdown

Shows how each feature moves the prediction from the baseline (expected value) to the
final predicted probability for one specific pitch. Change `row_id` to inspect different pitches.

```{r shap-waterfall, fig.width=10, fig.height=7}
# row_id = 1 is the first pitch in the filtered set; change as desired
waterfall_row <- 1

pitch_meta <- player_data |>
  slice(waterfall_row) |>
  select(game_date, pitch_name)

p_waterfall <- sv_waterfall(shap_obj, row_id = waterfall_row) +
  labs(title = paste0(target_player, "  |  Single Pitch SHAP Breakdown"),
       subtitle = paste0(pitch_meta$pitch_name, "  —  ", pitch_meta$game_date))

p_waterfall

ggsave(filename = file.path(output_dir,
         paste0(gsub(" ", "_", target_player), "_", target_hand, "_",
                pitch_type, "_", outcome_type, "_SHAP_waterfall_pitch", waterfall_row, ".png")),
       plot = p_waterfall, width = 10, height = 7, dpi = 300)
```

### Waterfall — Average Across All Pitches

A single "typical pitch" view: the mean SHAP values across all of the player's pitches,
showing which features systematically push their prediction up or down.

```{r shap-waterfall-mean, fig.width=10, fig.height=7}
p_waterfall_mean <- sv_waterfall(shap_obj, row_id = 1:nrow(X_player)) +
  labs(title = paste0(target_player, "  |  Mean SHAP Breakdown"),
       subtitle = paste0("Averaged across all ", nrow(X_player), " ", pitch_type, " pitches"))

p_waterfall_mean

ggsave(filename = file.path(output_dir,
         paste0(gsub(" ", "_", target_player), "_", target_hand, "_",
                pitch_type, "_", outcome_type, "_SHAP_waterfall_mean.png")),
       plot = p_waterfall_mean, width = 10, height = 7, dpi = 300)
```
